Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-21 ms.tex
cyclic

In this paper we instead focus on time-independent models, in cases when the nuclear gas receives sufficient heating to render radiative cooling negligible.  This approach allows us to systematically explore the relevant parameter
space and to derive analytic expressions that prove useful in
determining whether cooling instabilities manifest across the expected
range of galaxy properties, or whether other (non-AGN) forms of
feedback can produce a prolonged state of steady, thermally stable
accretion.  Even if cooling instabilities develop on galactic scales
over longer $\sim$ Gyr timescales, a quasi steady-state may exist
between these inflow events on smaller radial scales comparable to the
sphere of influence.  We are thus motivated to explore whether such
steady flows are generally expected within the nucleus, or
whether in most galaxies the nuclear gas content is always building
towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ usually resides near the nominal Bondi
radius).  When heating is sufficiently weak, however, the stagnation
radius may move to much larger radii or not exist at all, in which
case nearly the entire ISM of the galaxy is flowing towards the SMBH.
Here we assess under what conditions a stagnation
radius exists, a question intimately tied to that of cooling
instability.

@@

\citet{Sharma+2007} calculate the radiative efficiency, $\epsilon$,
based on MHD shearing box simulations of collisionless plasmas. For
accretion rates $10^{-4} \gsim \eddr \lsim 10^{-2}$, the radiative
efficiency is $\sim 2.6\%$ {\bf AG described in more
  detail?}. Additionally, a large fraction of the $\dot{M}$ on large scales
will be driven out in outflows as the gas has some non-zero angular
momentum and a centrifugal barrier will stop it from accreting
\citet{Li+2013}. In the limit weak cooling appropriate for low
accretion rates the accretion rate on small scales will be of order
$\alpha \Mdot$. Here we adopt $\alpha=0.1$ so that our overall
efficiency is 0.26\%

@@

%Many of the upper
%limits on $L_{\rm X}$ at low $M_{\bullet}$ may in fact correspond to a
%separate ``thermally stable" population.  This possible dichotomy
%calls into question the use of simple extrapolations of power-law fits
%to the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to
%constrain the occupation fraction of black holes in low mass galaxies
%(\citealt{Miller+15}).

\subsection{Steady Accretion versus Cyclic Outbursts}
\label{sec:cycle}

It is also possible that many of the \citet{Miller+15} sample of
galactic nuclei are not accreting in steady state.  The 
X-ray efficiency assumed in the right panel of Figure \ref{fig:miller}
represents a maximum value, which does not for instance account for
suppression of the SMBH accretion rate due to outflows from the
accretion disk on small radial scales (e.g.,
\citealt{Blandford&Begelman99}).  Furthermore, our SFH-averaged
accretion rates are derived using analytic expressions for the
stagnation radius that may underpredict the true accretion rate if
$\zeta < \zeta_c$, as may occur in massive core galaxies with large
break radii.

If the actual radiative efficiency is lower than our
fiducial assumption, then many of the X-ray detections would lie above
the predictions for stable accretion, i.e., $L_X \gtrsim \epsilon_X \dot{M}_{\rm TI}$ (yellow line).  In this case the X-ray detected galaxies experience
less heating than average galaxies of the same mass, and may not be
stably accreting at all.  This is particularly plausible because the
\citet{Miller+15} sample includes only early-type galaxies, with older
stellar populations than the average for their black hole mass.

@@

If present, thermal instabilities may cause the CNM to undergo cyclic
oscillations of rapid cooling and high accretion rates, followed by
quiescent periods once gas has been consumed and/or AGN feedback
becomes effective.  Such cycles are seen in numerical simulations of
elliptical galaxies on large scales and long timescales
(\citealt{Ciotti+10}).  Evidence for such periodic outbursts includes
observations showing that a fraction of early-type galaxies in the
local universe have undergone recent ($< 1-2$ Gyr old) star formation
episodes (\citealt{Donas+07}).

@@

%This could help explain why core galaxies show a wide
%range of nuclear radio luminosities (suggesting cyclic
%accretion episodes), while cusp galaxies are confined below a
%threshold (e.g., \citealt{Bender+89, Pellegrini99,
 % Capetti&Balmaverde05}), consistent with thermally-stable,
%steady-state picture. ({\bf AG: This picture may be dubious...Review
 % the literature but I think that the higher upper bound for cores is
  %just due to a mass selection effect.  BDM: Balmaverde
 % disagrees.})

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$
in Fig \ref{fig:miller} are indeed in a state of thermally-unstable
cyclic behavior, then a subset of the galaxies with X-ray upper limits
may contribute to a separate population of thermally stable, steadily
accreting nuclei.  Such a population could include SgrA*, which has a
much lower X-ray luminosity for its black hole mass than predicted by
the \citet{Miller+15} trend.  A potentially bimodal population of
steady and cyclic galactic nuclei calls into question the physical
justification of using simple extrapolations of power-law fits to the
$L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to
constrain the occupation fraction of SMBHs in the nuclei of low mass
galaxies.

%The cyclic model predicts that high $L_{X}$ nuclei should be more
%common in galaxies with younger stellar ages, a result which appears
%in tension with the finding by \citet{Pellegrini10} that galaxies with
%younger stellar populations are observed to host less X-ray luminous
%nuclei, at least for stellar ages $<3$ Gyr.  An alternative
%explanation for this observation within a thermally stable,
%steady-state accretion picture is that young stars provide higher
%stellar wind heating, reducing the mass accretion rate and hence
%nuclear X-ray luminosity.  {\bf AG Challenge: \citet{HeckmanBest:2014a}
 % find that ${\rm L_{[OIII]}}$ (a proxy for the bolometric luminosity)
  %${\rm L_{[OIII]}/M_\star}$ (a proxy for the Eddington ratio)
  %increases with decreasing 4000 $\AA$ break index (decreasing stellar
 % age) (see their Figure 14). This suggests that the accretion rate is
  %actually increasing towards younger stellar ages. At young ages
  %there actually seems to be plenty of gas around, which may be
  %obscuring the x-rays and causing an anamolously low measurement of
  %$L_X$.

  %For a starburst the black hole accretion rate lags the star
  %formation rate by 200 Myr (\citet{WildHeckman+:2010a} Figure
 % 9). This could be due to SNe feed back suppressing the black hole
  %accretion rate (massive star winds do the same thing by themselves,
 % but they cannot be sustained for 200 Myr. 
%}

@@

%The implications of our model for the CNM profiles of quiescent galaxies on the diversity of radio emission from TDE
%jets will be further developed in future work.

>>>>

Between outbursts \citealt{Ciotti+10} find that galactic nuclei exist
in a quasi-stationary thermal stable state. We focus on this region of
parameter space calculating time-independent models, in which the
nuclear gas receives sufficient heating to render radiative cooling
negligible.  This approach allows us to systematically explore the
relevant parameter space and to derive analytic expressions that prove
useful in determining under what conditions cooling instabilities
manifest across the expected range of galaxy properties, or whether
other (non-AGN) forms of feedback can produce a prolonged state of
steady, thermally stable accretion.  Even if cooling instabilities
develop on galactic scales over longer $\sim$ Gyr timescales, a quasi
steady-state may exist between these inflow events on smaller radial
scales comparable to the sphere of influence.  We are thus motivated
to explore whether such steady flows are generally expected within the
nucleus, or whether in most galaxies the nuclear gas content is always
building towards instability.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ usually resides near the nominal Bondi
radius).  When heating is sufficiently weak, however, the stagnation
radius may move to much larger radii or not exist at all, in which
case nearly the entire ISM of the galaxy is flowing towards the SMBH.
Here we assess under what conditions a stagnation
radius exists, a question intimately tied to that of cooling
instability. {\bf AG this section still needs work...}

@@

\citet{Sharma+2007} calculate the radiative efficiency, $\epsilon$,
based on MHD shearing box simulations of collisionless plasmas. For
accretion rates $10^{-4} \lsim \eddr \lsim 10^{-2}$, the radiative
efficiency is $\sim 2.6\%$ {\bf AG described in more
  detail?}. Additionally, a large fraction of the $\dot{M}$ on large scales
will be driven out in outflows as the gas has some non-zero angular
momentum and a centrifugal barrier will stop it from accreting
\citet{Li+2013}. In the limit weak cooling appropriate for low
accretion rates the accretion rate on small scales will be of order
$\alpha \Mdot$. Here we adopt $\alpha=0.1$ so that our overall
efficiency is 0.26\%

@@

\subsection{Steady Accretion versus outbursts}
\label{sec:cycle}

It is also possible that many of the \citet{Miller+15} sample of
galactic nuclei are not accreting in steady state. Within our steady
state solutions gas outside the stagnation radius $\rs$ is being blown
out in wind. However, for very massive galaxies ($\Mbh\gsim 10^8
\Msun$) the energy injected into the gas is not enough to truly blow
the gas out of the potential well of the galaxy. This gas is likely to
build up on the outskirts of the galaxy until a thermal instability
develops, causing the CNM to undergo cyclic oscillations of rapid
cooling and high accretion rates, followed by quiescent periods once
gas has been consumed and/or AGN feedback becomes effective. Such
cycles are seen in numerical simulations of elliptical galaxies on
large scales and long timescales (\citealt{Ciotti+10}).  Evidence for
such periodic outbursts includes observations showing that a fraction
of early-type galaxies in the local universe have undergone recent ($<
1-2$ Gyr old) star formation episodes (\citealt{Donas+07}).

Furthermore, our SFH-averaged accretion rates are derived using
analytic expressions for the stagnation radius that may underpredict
the true accretion rate if $\zeta < \zeta_c$, as may occur in massive
core galaxies with large break radii.

If the actual radiative efficiency is lower than our fiducial
assumption, then many of the X-ray detections would lie above the
predictions for stable accretion, i.e., $L_X \gtrsim \epsilon_X
\dot{M}_{\rm TI}$ (yellow line).  For example, we assume 10\% 
of the mass inflow rate is drive out in small scale outflows, but this
fraction could be considerably large. In this case the X-ray detected
galaxies experience less heating than average galaxies of the same
mass, and may not be stably accreting at all.  This is particularly
plausible because the \citet{Miller+15} sample includes only
early-type galaxies, with older stellar populations than the average
for their black hole mass.

@@

@@

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$
in Fig \ref{fig:miller} are indeed in a state of thermally-unstable
outburst, then a subset of the galaxies with X-ray upper limits
may contribute to a separate population of thermally stable, steadily
accreting nuclei.  Such a population could include SgrA*, which has a
much lower X-ray luminosity for its black hole mass than predicted by
the \citet{Miller+15} trend.  A potentially bimodal population of
steady and cyclic galactic nuclei calls into question the physical
justification of using simple extrapolations of power-law fits to the
$L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to
constrain the occupation fraction of SMBHs in the nuclei of low mass
galaxies.

@@
